Looking for a Signal in the Noise: Revisiting Obesity and the microbiome

### 2016-03-05 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
added ZScoreLogBF

@@

@@

@@

@@

@@

@@

@@

@@

@@

#Load in BMI data
data1 <- read.csv("combined.H.corr.data.csv")
#Load in Turnbaugh data
turn.data <- read.csv("Turnbaugh.normalized.csv")

#Turnbaugh STD = 0.4818751
turn.ZScore.H <- turn.data$H.corr/0.4818751

#Amish STD = 0.5932651
keepamish <- which(data1$dataset == "Amish")
amishdata <- data1[keepamish, ]
amish.ZScore.H <- amishdata$H.corr/0.5932651

#MetaHit-Den STD = 0.4038558
keepmetahit <- which(data1$dataset == "MetaHit")
metahitdata <- data1[keepmetahit, ]
metahit.ZScore.H <- metahitdata$H.corr/0.4038558

#Korean STD = 0.5400615
keepKorean <- which(data1$dataset == "Korean")
Koreandata <- data1[keepKorean, ]
Korean.ZScore.H <- Koreandata$H.corr/0.5400615

#HMP STD = 0.6774514
keepHMP <- which(data1$dataset == "HMP")
hmpdata <- data1[keepHMP, ]
HMP.ZScore.H <- hmpdata$H.corr/0.6774514

#CCHC STD = 0.5784522
keepCCHC <- which(data1$dataset == "CCHC")
CCHCdata <- data1[keepCCHC, ]
CCHC.ZScore.H <- CCHCdata$H.corr/0.5784522

#GLNE07 STD = 0.4663871
keepGLNE <- which(data1$dataset == "GLNE")
GLNEdata <- data1[keepGLNE, ]
GLNE.ZScore.H <- GLNEdata$H.corr/0.4663871

#Escobar STD = 0.5378935
keepescobar <- which(data1$dataset == "Escobar")
escobardata <- data1[keepescobar, ]
Escobar.ZScore.H <- escobardata$H.corr/0.5378935

#COMBO STD = 0.6870763
keepCOMBO <- which(data1$dataset == "COMBO")
COMBOdata <- data1[keepCOMBO, ]
COMBO.ZScore.H <- COMBOdata$H.corr/0.6870763

#COMBO STD = 0.3489582
keeptwinsUK <- which(data1$dataset == "twinsUK")
twinsUKdata <- data1[keeptwinsUK, ]
twinsUK.ZScore.H <- twinsUKdata$H.corr/0.3489582

#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit, COMBO, twinsUK
HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H, amish.ZScore.H)
data1 <- cbind(HZScore, data1)

turn.data <- cbind(turn.ZScore.H, turn.data)
rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)

@@

#Create group analysis table to harness full data set
#For ZScore normalization
d1.obese <- data1$obese
t.obese <- as.character(turn.data$obese)

d1.bmi.class <- data1$BMI.class
t.bmi.class <- as.character(turn.data$BMI.category)

d1.H.Zscore <- data1$HZScore
t.H.Zscore <- turn.data$turn.ZScore.H

@@

setwd("C:/Users/marc/Desktop/obesity2/BF.normalization")

@@

#Load in data
data2 <- read.csv("combined.BF.corr.csv")

#Create BMI groups
data2$BMI.class[data2$bmi<=24] <- "Lean"
data2$BMI.class[data2$bmi>24 & data2$bmi<30] <- "Overweight"
data2$BMI.class[data2$bmi>=30] <- "Obese"

#Create Obese Yes/No groups
data2$obese[data2$BMI.class=="Lean" | data2$BMI.class=="Overweight"] <- "No"
data2$obese[data2$BMI.class=="Obese"] <- "Yes"

##Run Zscore normalized Bacter and Firm overall analysis
#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
d1.zbf.corr <- data2$ZBF
t.zbf.corr <- turn.data$ZBF

#Bacteroidetes/Firmicutes Datasets
comb.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr))
group.data.test <- cbind(comb.zbf.corr, group.data.test)
colnames(group.data.test)[1] <- "zbf"

@@

# Section C
bf1 <- ggplot(group.data.test, aes(obese, zbf))
bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized B/F Ratio")
bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
bf6 <- bf5 + theme(axis.line=element_line(colour="black"),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank(),
                   panel.border = element_blank(),
                   panel.background = element_blank())
bf7 <- bf6 + ggtitle('C') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section D
qbf1 <- ggplot(group.data.test, aes(BMI.class, zbf))
qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized B/F Ratio")
qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
qbf6 <- qbf5 + theme(axis.line=element_line(colour="black"),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     panel.border = element_blank(),
                     panel.background = element_blank())
qbf7 <- qbf6 + ggtitle('D') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

@@

**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity. There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = 0.005992). **B)** ZScore Normalized Shannon Diversity by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = 0.018) but no difference between lean and obese group (P-value = 0.081). **C)** ZScore Normalized B/F Ratio and Obesity. There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = 0.456). **D)** ZScore Normalized B/F Ratio by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = 0.398).

>>>>

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

BaxterZH <- scale(H)
BaxterZLogBF <- scale(log(BFratio))
BaxterBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

RossZH <- scale(H)
RossZLogBF <- scale(log(BFratio))
RossBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

GoodrichZH <- scale(H)
GoodrichZLogBF <- scale(log(BFratio))
GoodrichBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

EscobarZH <- scale(H)
EscobarZLogBF <- scale(log(BFratio))
EscobarBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

ZupancicZH <- scale(H)
ZupancicZLogBF <- scale(log(BFratio))
ZupancicBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

HMPZH <- scale(H)
HMPZLogBF <- scale(log(BFratio))
HMPBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

WuZH <- scale(H)
WuZLogBF <- scale(log(BFratio))
WuBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

ArumugamZH <- scale(H)
ArumugamZLogBF <- scale(log(BFratio))
ArumugamBMI <- bmi

@@

###########################################################################
############ ZScore Data Preperation ###################################
###########################################################################

TurnbaughZH <- scale(H)
TurnbaughZLogBF <- scale(log(BFratio))
TurnbaughObese <- as.character(obese)
TurnbaughBMICat <- as.character(bmi)

@@

#data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic

ZH <- c(EscobarZH, BaxterZH, RossZH, HMPZH, ArumugamZH, WuZH, GoodrichZH, ZupancicZH)
bmi <- c(EscobarBMI, BaxterBMI, RossBMI, HMPBMI, ArumugamBMI, WuBMI, GoodrichBMI, ZupancicBMI)
ZlogBF <- c(EscobarZLogBF, BaxterZLogBF, RossZLogBF, HMPZLogBF, ArumugamZLogBF, WuZLogBF, GoodrichZLogBF, ZupancicZLogBF)

data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))

turn.comb.meta <- as.data.frame(cbind(TurnbaughBMICat, TurnbaughObese))
turn.comb.cont <- as.data.frame(cbind(TurnbaughZH, TurnbaughZLogBF))
turn.data <- cbind(turn.comb.cont, turn.comb.meta)
colnames(turn.data) <- c("ZH", "ZLogBF", "BMI.class", "obese")

@@

#Create group analysis table to harness full data set
#For ZScore normalization
d1.obese <- as.character(data1$obese)
t.obese <- as.character(turn.data$obese)

d1.bmi.class <- data1$BMI.class
t.bmi.class <- as.character(turn.data$BMI.class)

d1.H.Zscore <- data1$ZH
t.H.Zscore <- turn.data$ZH

@@

@@

##Run Zscore normalized Bacter and Firm overall analysis
#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
d1.lzbf.corr <- data1$ZlogBF
t.lzbf.corr <- turn.data$ZLogBF

#Bacteroidetes/Firmicutes Datasets
comb.lzbf.corr <- as.data.frame(c(d1.lzbf.corr, t.lzbf.corr))
group.data.test <- cbind(comb.lzbf.corr, group.data.test)
colnames(group.data.test)[1] <- "ZLogBF"

## Perform the statistical tests
combZHTtest <- t.test(H.Zscore ~ obese, data=group.data.test) #P-value = 0.006061
combZHANOVA <- anova(lm(H.Zscore ~ BMI.class, data=group.data.test)) #P-value = 0.0193
a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test)
combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
combZHTukeyFinal <- combZHTukey$BMI.class

combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.866
combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.106
a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
combZLogBFTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
combZLogBFTukeyFinal <- combZLogBFTukey$BMI.class

@@

# Section C
bf1 <- ggplot(group.data.test, aes(obese, ZLogBF))
bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized Log B/F Ratio")
bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
bf6 <- bf5 + theme(axis.line=element_line(colour="black"),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank(),
                   panel.border = element_blank(),
                   panel.background = element_blank())
bf7 <- bf6 + ggtitle('C') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

# Section D
qbf1 <- ggplot(group.data.test, aes(BMI.class, ZLogBF))
qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized Log B/F Ratio")
qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
qbf6 <- qbf5 + theme(axis.line=element_line(colour="black"),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     panel.border = element_blank(),
                     panel.background = element_blank())
qbf7 <- qbf6 + ggtitle('D') + theme(plot.title=element_text(hjust=0, face="bold", size=20))

@@

**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity. There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = `r signif(combZHTtest$p.value, 3)`). **B)** ZScore Normalized Shannon Diversity by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`). **C)** ZScore Normalized B/F Ratio and Obesity. There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`). **D)** ZScore Normalized B/F Ratio by BMI Group. Total n for lean, overweight, and obese groups are 563, 435, 518 respectively. There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = `r signif(combZLogBFANOVA[1, 5], 3)`).
